---
title: "Getting started with dbiutils"
description: >
  Execute parameterized DBI queries, manage short-lived connections, and load
  SQL statements from files.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with dbiutils}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`dbiutils` provides three small helpers that remain independent of the DBI
backend selected by an application:

- `db_query()` executes a query and returns a `data.table`;
- `with_db_connection()` scopes the lifecycle of a short-lived connection;
- `read_sql_file()` loads a SQL statement from an explicitly located file.

This vignette uses an in-memory SQLite database, but the same helpers work with
`RPostgres`, `odbc`, and other DBI-compatible backends.

## Manage a short-lived connection

A connection provider is a zero-argument function that creates a new
connection. `with_db_connection()` owns the returned connection and guarantees
that it is disconnected after the callback completes, including when the
callback raises an error.

```{r connection}
library(dbiutils)

sqlite_connection <- function() {
  DBI::dbConnect(RSQLite::SQLite(), ":memory:")
}

values <- with_db_connection(sqlite_connection, function(connection) {
  DBI::dbWriteTable(
    connection,
    "numbers",
    data.frame(value = 1:5)
  )

  db_query(
    connection,
    "SELECT value FROM numbers WHERE value > ? ORDER BY value",
    params = list(2)
  )
})

values
```

Do not use `with_db_connection()` when the provider returns a shared connection
whose lifecycle is managed elsewhere.

## Execute parameterized queries

`db_query()` delegates query execution and parameter binding to DBI, then
converts the result to a `data.table`. Placeholder syntax remains specific to
the selected backend.

```{r query}
with_db_connection(sqlite_connection, function(connection) {
  db_query(
    connection,
    "SELECT ? AS package, ? AS purpose",
    params = list("dbiutils", "reusable DBI helpers")
  )
})
```

Prefer parameter binding over constructing SQL by concatenating external
values.

## Load SQL from a file

Keeping larger statements in separate files makes them easier to review and
reuse. `read_sql_file()` expects the complete path and reads UTF-8 by default.

```{r sql-file}
sql_file <- tempfile(fileext = ".sql")
writeLines(
  c(
    "SELECT value",
    "FROM numbers",
    "WHERE value >= ?",
    "ORDER BY value"
  ),
  sql_file,
  useBytes = TRUE
)

sql <- read_sql_file(sql_file)
sql
```

Inside another R package, locate packaged SQL resources with `system.file()`:

```{r system-file, eval = FALSE}
sql_path <- system.file(
  "sql",
  "find_numbers.sql",
  package = "yourpackage",
  mustWork = TRUE
)

sql <- read_sql_file(sql_path)
```

## Backend responsibility

`dbiutils` imports `DBI` and `data.table` but intentionally imports no database
driver. The consuming application remains responsible for installing,
configuring, and authenticating its selected backend.
